Highly Available Transactions: Virtues and Limitations

### 2013-03-27 writeup/intro.tex
neilc comments on intro

Requirements for high availability, predictable low latency, and
simplicity of coordination protocols have led many recent distributed
database designers to renounce several guarantees provided by
traditional database systems. In particular, many modern systems forgo
providing ACID guarantees---which often provide single-system image
(SSI) programmability but suffer from unavailability and often
expensive coordination costs. The CAP Theorem, demonstrating the
incompatibility of ``always on'' operation with guarantees on data
item recency in the presence of unreliable networks, has led a large
number of systems to prioritize high availability and guaranteed low
latency, which are especially important in geo-replicated contexts. We
observe a dearth of systems providing general-purpose multi-operation
\textit{transactions} over multiple data items with useful semantics.

The relationship of the CAP Theorem and ACID semantics is poorly
understood. While serializability---the gold standard of SSI
transactions---is unavailable, databases have a long tradition of
providing weaker isolation and consistency guarantees. While these
weaker guarantees do not provide SSI for arbitrary general-purpose
read/write transactions, they often admit greater concurrency and
higher performance than comparable SSI algorithms. Accordingly, weak
isolation is widely employed, overwhelmingly as the default setting in
today's traditional ``ACID'' and up-and-coming ``NewSQL'' databases
and often as the maximum guarantees offered
(Section~\ref{sec:modernacid}). While the prevalence of these
configurations is encouraging, the current literature does not provide
guidance as to the operation of weakly consistent isolation guarantees
in a distributed context. Rather, we have a strong understanding of
weak isolation in the single-server context from which it
originated~\cite{adya, ansicritique, gray-isolation} as well as
several techniques for providing distributed
serializability~\cite{bernstein-concurrency} but we have little to no
insight into which ACID guarantees are compatible with Highly
Available Transactions (HATs), or can be provided with high
availability and low latency.

This work attempts to bridge the gap between ACID properties and
highly available systems. We analyze a range of existing ACID and
distributed consistency models to provide a rigorous taxonomization of
which models are achievable in a HAT system.  While several
guarantees, like serializability and Snapshot Isolation, are
unachievable due to reliance on preventing Lost Update and Write Skew
phenomena, many other ACID guarantees are HAT compliant, such as
providing Read Committed isolation, preventing Fuzzy Read phenomena,
and ensuring atomicity of transactional updates. We also attempt to
bridge the often distant worlds of distributed system consistency and
ACID database consistency by analyzing the relationship between
so-called session guarantees~\cite{sessionguarantees}, ``register''
semantics~\cite{herlihy-art}, and ACID properties. In doing so, we
introduce a hierarchy of highly available properties as, for example,
some guarantees require ``stickiness,'' or fate-sharing between
clients and database servers. We believe that this analysis is the
first that has been performed over such a broad range of ACID and
distributed consistency models.

While this model classification is useful, we also seek to understand
the implications for end-users of HAT systems. HATs provide high
availability and low latency but cannot offer all of the semantic
guarantees of unavailable systems---when is this trade-off worthwhile?
We attempt to objectively study these virtues and limitations of both
classic and HAT systems by performing an extensive survey of
practitioner accounts and research literature, performing experimental
analyses on modern cloud infrastructure, and analyzing representative
applications for their semantic requirements. While our results are
not definitive for all applications, they suggest that HATs offer a
one to three order of magnitude latency decrease compared to
traditional protocols and provide acceptable semantics for a wide
range of programs, particularly those with monotonic logic and
commutative updates~\cite{calm, blooml, crdt}. HAT systems can also provide
arbitrary foreign key constraints on multi-item updates and sometimes
provide uniqeness guarantees. However, particularly for programs with
non-monotonic logic, HATs may fall short, requiring more
coordination-intensive mechanisms. Concisely, we attempt to understand
\textit{what benefits do Highly Available Transactions offer and what
  is their cost?}

@@

\begin{itemize}
\item We present a model for high availability in a transactional
  environment, replica-based High Availability, Sticky High
  Availability, and Transactional High Availability.

\item We analyze the existing literature on ACID properties and
  replicated data semantics to show which properties are achievable in
  HAT systems. Achievable properties include transactional atomicity,
  variants of repeatable read isolation, and session guarantees. We
  also describe unavailable properties, like preventing Lost Update
  and Write Skew. We believe that this is the first such
  characterization in the literature.

\item We further motivate the study of HATs via the analysis of
  existing concurrency control algorithms, a case-study of running a
  transactional application with HATs, and via deployment of a HAT
  database prototype on public cloud infrastructure.
\end{itemize}

>>>>

Requirements for high availability, predictable low latency, and
simplicity of coordination protocols have led many recent distributed
database designers to renounce several guarantees provided by
traditional database systems~\cite{bigtable, pnuts, dynamo}. In
particular, many modern systems forgo providing ACID
guarantees---which often provide single-system image programmability
but suffer from unavailability and often expensive coordination
costs~\cite{davidson-survey}. The CAP Theorem~\cite{gilbert-cap},
demonstrating the incompatibility of ``always on'' operation with
guarantees on data item recency in the presence of unreliable
networks, has led a large number of systems to prioritize high
availability and guaranteed low latency, which are especially
important in geo-replicated contexts. We observe a dearth of highly
available systems providing general-purpose multi-operation
\textit{transactions} over multiple data items with useful semantics.

While serializability---the gold standard of traditional ACID
transactions---is not achievable with high availability for general
read-write transactions~\cite{davidson-survey}, databases have a long
tradition of providing weaker isolation and consistency
guarantees~\cite{adya, ansicritique, gray-isolation}. Due to their
concurrency and overall performance benefits, weak isolation models
are widely employed, overwhelmingly as the default setting in today's
traditional ``ACID'' and up-and-coming ``NewSQL'' databases and often
as the maximum guarantee offered (Section~\ref{sec:modernacid}). While
this weak isolation does not provide serializability for
general-purpose transactions, it is apparently strong enough to
deliver acceptable behavior to many application programmers and is
substantially stronger than the typical semantics of current highly
available systems. This raises the question: which of these semantics
can we provide while remaining highly available?

Unfortunately, the relationship of the CAP Theorem and ACID semantics
is poorly understood. We have a strong understanding of weak isolation
in the single-server context from which it originated~\cite{adya,
  ansicritique, gray-isolation} as well as several techniques for
providing distributed
serializability~\cite{bernstein-concurrency}. However, the current
literature does not provide guidance as to the operation of weakly
consistent transactional guarantees in a highly available distributed
context. Accordingly, in this work, we consider the problem of
providing Highly Available Transactions (HATs), or semantic guarantees
regarding multiple operations operating on multiple data items that
are achievable with high availability and low latency, albeit with
some cost to achievable semantics. While our preliminary results
indicate that some models are achievable~\cite{hat-hotos}, the
literature lacks a deep understanding of what models are available,
which are not, and when this matters.

This work attempts to bridge the gap between ACID properties and
highly available systems. We introduce a hierarchy of consistency
models and rigorously taxonomize their availability characteristics;
in doing so, we unify previous work on ACID properties~\cite{adya},
session guarantees~\cite{sessionguarantees}, and traditional
disributed ``register'' semantics~\cite{herlihy-art}. By identifying
the availability cost of a wide range of popular consistency
guarantees, this work allows developers to weigh the costs and
benefits of relying on these guarantees in their applications. While
several guarantees, like serializability and Snapshot Isolation, are
unachievable due to update conflict avoidance (preventing Lost Update
and Write Skew phenomena), many other ACID guarantees are HAT
compliant, such as providing Read Committed isolation, preventing
Fuzzy Read phenomena, and ensuring atomicity of transactional
updates. Several other guarantees, like reading one's writes, fall
into a new availability class---``sticky'' high availability---which
provides an intermediate ground between strict high availability and
majority availability.

This taxonomy allows us to weigh the costs and benefits of a wide
range of semantic guarantees, but which guarantees are worthwhile in
practice? We attempt to objectively study the virtues and limitations
of both classic and HAT systems by performing an extensive survey of
practitioner accounts and research literature, performing experimental
analyses on modern cloud infrastructure, and analyzing representative
applications for their semantic requirements. While our results are
not definitive for all applications, they suggest that HATs offer a
one to three order of magnitude latency decrease compared to
traditional protocols and provide acceptable semantics for a wide
range of programs, particularly those with monotonic logic and
commutative updates~\cite{calm, blooml, crdt}. HAT systems can also
provide arbitrary foreign key constraints on multi-item updates and
sometimes provide uniqeness guarantees. However, particularly for
programs with non-monotonic logic, HATs may fall short, requiring more
coordination-intensive mechanisms. Concisely, we attempt to understand
\textit{what benefits do Highly Available Transactions offer and what
  is their cost?}

@@

\begin{itemize}
\item We model high availability in a transactional environment,
  including traditional and sticky high availability.

\item We taxonomize ACID and distributed consistency properties
  according to their availability characteristics.

\item We analyze existing concurrency control algorithms, perform a
  case-study of an existing transactional application, and
  experimentally evaluate a HAT database prototype.
\end{itemize}

### 2013-03-27 writeup/motivation.tex
neilc comments on intro

Peter Deutsch begins his classic list of ``Fallacies of Distributed
Computing'' with two concerns fundamental to distributed database
systems: ``\textit{1.)}  The network is reliable. \textit{2.)} Latency
is zero''~\cite{fallacies-deutsch}. On a single server, communication
channels are robust and relatively fast, and few single-server systems
need to account for a possibility of partial system failure. However,
both of these assumptions are no longer valid in a distributed
setting. To extend Jim Gray's storage latency analogy, if memory is
1.5 hours from the CPU and disk is 2 years away, then, especially
across wide-area networks (WANs), remote server accesses can exceed 30
years, with the added concern that messages can be lost in transit or
delayed indefinitely by network
partitions~\cite{gray-rules}. Accordingly, ignoring network failures
and latency can ``cause big trouble and painful learning
experiences''~\cite{fallacies-deutsch}. As an example of an actual system rearchitecture,
less than one year after its announcement, Yahoo!'s PNUTS developers
explicitly added a support for weaker, highly available operation,
explaining that ``strict adherence [to strong consistency] leads to
difficult situations under network partitioning or server
failures...in many circumstances, applications need a relaxed
approach''~\cite{pnuts-update}.

@@

According to James Hamilton, Vice President and Distinguished Engineer
on the Amazon Web Services team, ``network partitions should be rare
but net gear continues to cause more issues than it
should''~\cite{hamilton-partitions}. Anecdotal evidence confirms
Hamilton's assertion. In April 2011, a network misconfiguration led to
a twelve-hour long series of outages across the Amazon EC2 and RDS
services~\cite{amazon-netpartition}. Subsequent misconfigurations and
partial failures such as another EC2 outage in October 2012 have led
to full site disruptions for popular web services like Reddit,
Foursquare, and Heroku~\cite{ec2-downsites}. At a global scale,
hardware failures---like the 2011 outages in Internet backbones in
North America and Europe due a router
bug~\cite{juniper-partition}---and misconfigurations---such as BGP
faults in 2008~\cite{pakistan-youtube} and
2010~\cite{research-experiment-partition}---can cause widespread
partitions. Many of our discussions with practitioners---especially
those operating on public cloud infrastructure---as well as reports
from large-scale operators like Google~\cite{dean-keynote} confirm
that partitions are a reality for service operators today.

@@

One of the most popularly celebrated impossibility results for
distributed systems is Brewer's CAP Theorem, which, as formally proven
by Gilbert and Lynch, states that it is impossible to maintain
linearizability (``C,'' or ``consistency''---albeit poorly named), or
the ability to read the last completed write, and replica availability
(``A'') in the presence of network partitions
(``P'')~\cite{gilbert-cap}. Brewer's Theorem has roots in decades of
distributed database designs~\cite{davidson-survey} and has had a
substantial impact on recent large-scale distributed systems design:
while CAP is narrowly scoped to the impossibility of building highly
available linearizable systems linearizability, it is often
interpreted to preclude a larger set of guarantees, including ACID, or
transactional atomicity, consistency, isolation and durability
guarantees, as evidenced by the lack of highly available systems
providing them.  As an example of an production database redesign
explicitly attributable to partition behavior, less than one year
after its announcement, Yahoo!'s PNUTS system developers explicitly
added a support for weaker, highly available operation, explaining
that ``strict adherence [to strong consistency] leads to difficult
situations under network partitioning or server failures...in many
circumstances, applications need a relaxed
approach''~\cite{pnuts-update}.

@@

\begin{table}
\begin{small}
\begin{tabular}{|l|c|c|}
\hline
Database & Default & Maximum\\\hline
Actian Ingres 10.0/10S & S & S\\
Aerospike & RC & RC\\
Akiban Persistit & SI & SI\\
Clustrix CLX 4100 & RR & ?\\
Greenplum 4.1 & RC & S \\
IBM DB2 10 for z/OS & CS & S\\
IBM Informix 11.50 & Depends & RR\\
MySQL 5.6 & RR & S \\
MemSQL 1b & RC & RC\\
MS SQL Server 2012 & RC & S \\
NuoDB & CR & CR\\
Oracle 11g & RC & SI\\
Oracle Berkeley DB & S & S\\
Oracle Berkeley DB JE & RR & S\\
Postgres 9.2.2 & RC & S\\
SAP HANA & RC & SI\\
ScaleDB 1.02 & RC & RC\\
VoltDB & S & S\\
\hline
\multicolumn{3}{|p{7cm}|}{{\footnotesize RC: read committed, RR: repeatable read, SI: snapshot isolation, S: serializability, CS: cursor stability, CR: consistent read}}\\\hline

\end{tabular}
\caption{Default and maximum isolation levels for ACID and NewSQL
  databases as of January 2013 (from
  \protect\cite{hat-hotos}).}\vspace{-1.5em}
\label{table:existing}
\end{small}
\end{table}

>>>>

Peter Deutsch begins his classic list of ``Fallacies of Distributed
Computing'' with two concerns fundamental to distributed database
systems: ``\textit{1.)}  The network is reliable. \textit{2.)} Latency
is zero''~\cite{fallacies-deutsch}. On a single server, communication
channels are robust and relatively fast, and few single-server systems
need to account for a possibility of partial system failure. However,
both of these assumptions are no longer valid in a distributed
setting. To extend Jim Gray's storage latency analogy, if memory is
1.5 hours from the CPU and disk is 2 years away, then, especially
across wide-area networks (WANs), remote server accesses can exceed 30
years, with the added concern that messages can be lost in transit or
delayed indefinitely by network
partitions~\cite{gray-rules}. Accordingly, ignoring network failures
and latency can ``cause big trouble and painful learning
experiences''~\cite{fallacies-deutsch}.

@@

According to James Hamilton, Vice President and Distinguished Engineer
on the Amazon Web Services team, ``network partitions should be rare
but net gear continues to cause more issues than it
should''~\cite{hamilton-partitions}. Anecdotal evidence confirms
Hamilton's assertion. In April 2011, a network misconfiguration led to
a twelve-hour long series of outages across the Amazon EC2 and RDS
services~\cite{amazon-netpartition}. Subsequent misconfigurations and
partial failures such as another EC2 outage in October 2012 have led
to full site disruptions for popular web services like Reddit,
Foursquare, and Heroku~\cite{ec2-downsites}. At a global scale,
hardware failures---like the 2011 outages in Internet backbones in
North America and Europe due a router
bug~\cite{juniper-partition}---and misconfigurations---such as BGP
faults in 2008~\cite{pakistan-youtube} and
2010~\cite{research-experiment-partition}---can cause widespread
partitions.

Many of our discussions with practitioners---especially those
operating on public cloud infrastructure---as well as reports from
large-scale operators like Google~\cite{dean-keynote} confirm that
partition management is an important consideration service operators
today. As an example of an actual system rearchitecture, less than one
year after its announcement, Yahoo!'s PNUTS developers explicitly
added a support for weaker, highly available operation. The engineers
explained that ``strict adherence [to strong consistency] leads to
difficult situations under network partitioning or server
failures...in many circumstances, applications need a relaxed
approach''~\cite{pnuts-update}.

@@

One of the most popularly celebrated impossibility results for
distributed systems is Brewer's CAP Theorem, which, as formally proven
by Gilbert and Lynch, states that it is impossible to maintain
linearizability (``C,'' or ``consistency''---albeit poorly named), or
the ability to read the last completed write, and replica availability
(``A'') in the presence of network partitions
(``P'')~\cite{gilbert-cap}. Brewer's Theorem has roots in decades of
distributed database designs~\cite{davidson-survey} and has had a
substantial impact on recent large-scale distributed systems design:
while CAP is narrowly scoped to the impossibility of building highly
available linearizable systems linearizability, it is often
interpreted to preclude a larger set of guarantees, including ACID, or
transactional atomicity, consistency, isolation and durability
guarantees, as evidenced by the lack of highly available systems
providing them.

@@

\begin{table}
\begin{center}
\begin{small}
\begin{tabular}{|l|c|c|}
\hline
Database & Default & Maximum\\\hline
Actian Ingres 10.0/10S & S & S\\
Aerospike & RC & RC\\
Akiban Persistit & SI & SI\\
Clustrix CLX 4100 & RR & ?\\
Greenplum 4.1 & RC & S \\
IBM DB2 10 for z/OS & CS & S\\
IBM Informix 11.50 & Depends & RR\\
MySQL 5.6 & RR & S \\
MemSQL 1b & RC & RC\\
MS SQL Server 2012 & RC & S \\
NuoDB & CR & CR\\
Oracle 11g & RC & SI\\
Oracle Berkeley DB & S & S\\
Oracle Berkeley DB JE & RR & S\\
Postgres 9.2.2 & RC & S\\
SAP HANA & RC & SI\\
ScaleDB 1.02 & RC & RC\\
VoltDB & S & S\\
\hline
\multicolumn{3}{|p{7cm}|}{{\footnotesize RC: read committed, RR: repeatable read, SI: snapshot isolation, S: serializability, CS: cursor stability, CR: consistent read}}\\\hline

\end{tabular}
\caption{Default and maximum isolation levels for ACID and NewSQL
  databases as of January 2013 (from
  \protect\cite{hat-hotos}).}\vspace{-1.5em}
\label{table:existing}
\end{small}
\end{center}
\end{table}
